DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-12-19 2015CompleteDerivation/completeDerivation.tex
Reviewed the first two sections

While some branches of physics (e.g. Newtonian mechanics, thermodynamics, special relativity) are derived from physical assumptions (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. It is the case for classical Lagrangian mechanics, where we start by assuming the existence of a Lagrangian function whose path integral is minimized along the motion. It is the case for classical Hamiltonian mechanics, where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations. It is the case for quantum mechanics, where the state space is a complex inner product vector space and the state evolves according to Schroedinger's equations. That is, at the very core of many of our physical theories we have mathematical definitions instead of physical concepts.

While we have grown accustomed to this state of affair, it is nonetheless troublesome as it has problematic consequences. Suppose you have a system in front of you, how can you tell whether it's Hamiltonian? Or classical? We know that these theories have different range of applicability. To describe a cannonball we should be using classical mechanics while for an electron we should be using quantum mechanics. Not all Hamiltonians admit a corresponding Lagrangian. How do we pick?

If all we have are mathematical definitions, the only answer is that it is a Hamiltonian system if it obeys Hamiltonian mechanics. While many are indeed satisfied by such answer, one has to admit that it begs the question. We do not have a clear idea of what an Hamiltonian system is and how it different from a Lagrangian one. Or what makes a system classical and how it is different form a quantum system.

We started this work with the unpretentious goal of exploring these questions. The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such dictionary was complete, we had to make sure that all the relevant mathematical framework could be recovered. This lead to the mission of re-deriving as much as possible starting from physical assumptions.

This approach pushed us further and further back to the basic definitions. For example, if we truly wanted to justify why the state space of Hamiltonian mechanics is charted by conjugate variables $q^i$ and $p_i$, we had to justify why it was a manifold as well, which in turn meant we needed to justify why it was a topological space. But when we did find physical concepts that mapped to these more basic mathematical definitions, we realized that this link was valid for a far wider range of cases: the concepts were more general than we expected.

To our surprise, this pushed us to bring in other areas that were not within our horizon. The same link that allowed to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed to justify Lagrangian mechanics also gave us constrained the motion to free particles under scalar/vector potential forces. Since this happened repeatedly, it lead to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but a fundamental framework to understand and link different parts of basic science and mathematics.

%It felt like, when trying to assemble a jigsaw puzzle, you realize that the reason you can't find a piece is because you assembled the previous parts incorrectly and you end with a larger part that fits together a lot better.

These unexpected connections lead to the realization that sometimes different mathematical concepts are expression of the same physical concept. That is: by putting physics first we can achieve a sort of unification of ideas that is not possible by math alone.

The goal of this particular work is to report on this progress. The claim is that the whole architectural framework is solid enough to be convinced that it is possible to derive classical/quantum Hamiltonian/Lagrangian mechanics starting from a few physical principle. This is not just an exercise in style: this approach allows to make intuitive connections between different areas of physics and mathematics.

The claim is that, starting from a few simple physical assumptions, we can derive classical/quantum Hamiltonian/Lagrangian mechanics. The claim is not that we have achieved perfection and that every proposition and justification cannot be improved. The claim is, simply, that it \emph{can} be done, and that doing so provides better insight and intuition that clarifies the physics being described and that allows to strengthen connections between different areas of math and physics. We will see that sometimes different mathematical concepts are the expression of one physical concept. Therefore, by putting physics, first we can achieve a sort of unification of ideas that is not possible by math alone.

While trying to re-derive classical, quantum and (to some extent) relativistic mechanics may seem too broad a scope, it is necessary to ensure that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum system are Hamiltonian systems, so clearly we haven't understood what makes a system Hamiltonian. The ability to apply the same physical concepts far and wide is what gives us confidence that these are truly general ideas and they capture the physical aspects we are describing.

As brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution will give us a dynamical system: a state and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Additionally assuming kinematic equivalence, trajectories in space tell us everything about the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumption. Alternatively, as the second assumption we can take irreducibility, studying the whole systems tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

In section \ref{sec:fundamental_model} we will describe in general what it means to study a physical system and how the assumption of determinism and reversibility plays into it. We will develop an abstract conceptual model that we will use when discussing the later assumptions.

In section \ref{sec:dynamical_systems} we will show how the physical concept of physical distinguishability maps to the mathematical concept of topological spaces. Deterministic and reversible is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

In section \ref{sec:reducibility} we will introduce the notion of a classical material, one for which we assume infinitesimal reducibility. The state of an arbitrary amount of material is a distribution over the states of the infinitesimal parts, the classical particles. As such distribution must be invariant under the arbitrary choice of units, for each variable $q^i$ that defines a unit there is a corresponding variable $k^i$ defined on the inverse of such unit. Classical particle states can be identified by conjugate pairs of state variable $(q^i, p_i)$ (i.e. a point in the cotangent bundle $T^*\mathcal{Q}$) and the number of possible initial conditions for each pair is given by $dq^i \wedge dp_i$ (i.e. the canonical symplectic form $\omega$). During evolution, the distribution must be mapped point by point which leads to Hamiltonian mechanics (i.e. a symplectomorphism on $T^*\mathcal{Q}$).

In section \ref{sec:relativity} we will show that a relativistic version of Hamilton's equations arises naturally when we consider time dependent laws of evolution for distributions of classical material. This also gives the notion of deterministic and reversible evolution a characterization that is independent of any transformation that mixes time and state variables $(q^i, p_i)$.

In section \ref{sec:Lagrangian} we will introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variable $(x^i, u^i)$ (i.e. position and velocity). This link will allow us to express the number of possible initial conditions for d.o.f. in terms of $dx^i$ and $du^i$ (i.e. a metric tensor $g$). Such expression allows to constrain the Hamiltonian to the one for massive particles under scalar potential forces. Such assumption also allows for a Lagrangian formulation.

@@

In section \ref{sec:irreducibility} we will introduce the notion of an irreducible material, one for which studying the whole distribution does not tell us anything about the motion of its infinitesimal parts, the fragment. A unit amount of such material will correspond to a quantum particle. As this means that the states of such material are invariant under permutations of the configurations for the fragments, we will show that the state space of an irreducible material is a complex vector space. As we will still be able to tell how similar or different two states are, this will gives an inner product to the state space. Such comparisons will be preserved by deterministic and reversible evolution which leads to Schroedinger equation (i.e. a unitary transformation).

@@

The work is organized into:
\begin{description}
  \item[Assumptions] these characterize the physical system we are studying and constitute the premise of our discussion. A \textbf{rationale} follows each assumption, which uses physical and sometimes philosophical arguments to motivate why (or why not) such an assumption makes sense (in a particular case).
  \item[Propositions] these characterize the mathematical properties of the physical concepts we are studying. A \textbf{justification} follows each proposition to show why such characterization follows from the physical properties of the system. While the style of a justification is formal, it is not a pure mathematical proof as it necessarily has to mix physical and mathematical arguments. Before or after each proposition there is a \textbf{discussion} that conceptually explains the motivation and the result in a less formal, more approachable way.

  Justifications and discussions are designed to be independent to allow two separate types of reading style: one focused on higher level concepts and one focused on making sure the details follow. This inevitably leads to some repetition.

  \item[Theorems] these are pure mathematical statements that are used to simplify propositions. A mathematical \textbf{proof} follows each proposition. No mathematical breakthrough should be expected as we mostly use well known results from different fields. Theorems are grouped in the appendix.
\end{description}
This should allow readers with different focus and background to concentrate on the parts that are more keen on.

>>>>

While some branches of physics (e.g. Newtonian mechanics, thermodynamics, special relativity) are derived from physical principles (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. It is the case for classical Lagrangian mechanics, where we start by assuming the existence of a Lagrangian function whose path integral is minimized along the motion. It is the case for classical Hamiltonian mechanics, where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations. It is the case for quantum mechanics, where the state space is a complex inner product vector space and the state evolves according to Schroedinger's equations. That is, at the very core of some of our physical theories we have mathematical definitions instead of physical ideas.

If we ask ``when is a particular physical system suitably described by Hamiltonian (or by classical) mechanics?" the only answer we have is ``when it fits the mathematical framework". While many seem to be content with this answer (e.g. ``you have to start from somewhere, it may as well be the math which at least is precise"), it begs the question: it does not tell us how and what physical properties of the system under study are being captured.

The fundamental problem is that mathematical structures are not enough to characterize the physical system being described. For example, the same math of linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems. The mathematical treatment is the same because it captures aspects of the description that are common to all cases. But for the same very reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistic/computer science/phylosophy: math captures the syntax (the relationships between the objects) but does not capture the semantic (the meaning of each object).

We started this work with the unpretentious goal of exploring simple questions such as the following. What do these mathematical symbols used in classical/quantum Hamiltonian represent? What is the physical meaning attached to it? Why is the state of a quantum system described by a wave function (i.e. a vector in a complex inner product space), while the state of a classical hamiltonian system is described by the three pairs $(q^i, p_i)$ (i.e. a point in a cotangent bundle)? The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such dictionary was complete, we realized we had to make sure that the whole relevant mathematical framework could be recovered.

This redefined our mission to re-deriving as much as possible starting from physical assumptions. It also forced us to push further and further back toward the basic definitions. For example, if we truly wanted to justify why the state space of Hamiltonian mechanics is charted by conjugate variables $q^i$ and $p_i$, we had to justify why it was a manifold as well, which in turn meant we needed to justify why it was a topological space. But when we did find physical concepts that mapped to these more basic mathematical definitions, we realized that this link was valid for a far wider range of cases: the concepts were more general than we anticipated. For example, we realized that the mathematical notion of a topology maps incredibly well to the idea of physical distinguishability, but states of Hamiltonian system are not the only thing that are physically distinguishable.

To our surprise, this naturally expanded the boundary of the work. The same link that allowed to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed to justify Lagrangian mechanics also onstrained the motion to that of free particles under scalar/vector potential forces. We also realized that sometimes different mathematical concepts could be justified from the same physical starting point. That is: by putting physics first we could achieve a sort of unification of ideas that was not practicable by math alone.

The success of the approach lead us to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but instead a better framework to understand and link different parts of basic science and mathematics. After many years of work and different incremental iterations we think it is time to checkpoint and report the progress.

The goal of this paper is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian mechanics starting from a few physical principle. We do \emph{not} claim that we have perfectly achieved this result, with no room from improvement. We do claim that the whole architectural framework is solid enough to be convinced that this is indeed possible, and that it is a worthwhile endeavor as it allows to make intuitive connections between different areas of physics and mathematics. That is: it \emph{can} and \emph{should} be done.

While trying to re-derive classical, quantum and (to some extent) relativistic mechanics may seem too broad a scope, it is necessary to ensure that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum system are Hamiltonian systems, so clearly we haven't understood what makes a system Hamiltonian. The ability to apply the same physical ideas far and wide is what gives us confidence that these are truly general concept that capture the physical aspects we are describing. The breath itself is what makes the work interesting and more than a curiosity.

As a brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution will give us a dynamical system: a state and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Additionally assuming kinematic equivalence, trajectories in space tell us everything about the state of the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumption. Alternatively, as the second assumption we can take irreducibility, studying the whole systems tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

As a more detailed overview: in section \ref{sec:fundamental_model} we will describe in general what it means to study a physical system and how the assumption of determinism and reversibility plays into it. We will develop an abstract conceptual model that we will use when discussing the later assumptions.

In section \ref{sec:dynamical_systems} we will show how the physical concept of physical distinguishability maps to the mathematical concept of topological spaces. Deterministic and reversible evolution is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

In section \ref{sec:reducibility} we will introduce the notion of a classical material, one for which we assume infinitesimal reducibility. The state of an arbitrary amount of material is a distribution over the states of the infinitesimal parts, the classical particles. As such distribution must be invariant under the arbitrary choice of units, we'll see that for each variable $q^i$ that defines a unit there is a corresponding variable $k^i$ defined on the inverse of such unit. Classical particle states can be identified by conjugate pairs of state variable $(q^i, p_i)$ (i.e. a point in the cotangent bundle $T^*\mathcal{Q}$) and the number of possible initial conditions for each pair is given by $dq^i \wedge dp_i$ (i.e. the canonical symplectic form $\omega$). During evolution, the distribution (and marginal distributions) must be mapped point by point which leads to Hamiltonian mechanics (i.e. a symplectomorphism on $T^*\mathcal{Q}$).

In section \ref{sec:relativity} we will show that a relativistic version of Hamilton's equations arises naturally when we consider time dependent laws of evolution for distributions of classical material. This also gives the notion of deterministic and reversible evolution a characterization that is independent of any transformation that mixes time and state variables $(q^i, p_i)$. It will also give us a classical equivalent for anti-particles states.

In section \ref{sec:Lagrangian} we will introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variable $(x^i, u^i)$ (i.e. position and velocity). This link will allow us to express the number of possible initial conditions for d.o.f. in terms of $dx^i$ and $du^i$ (leading to the metric tensor $g$). Such expression also allows to constrain the Hamiltonian to the one for massive particles under scalar potential forces and an equivalent Lagrangian formulation.

@@

In section \ref{sec:irreducibility} we will introduce the notion of an irreducible material, one for which studying the whole distribution does not tell us anything about the motion of its infinitesimal parts, the fragments. A unit amount will correspond to a quantum particle. As this means that the states of such material are invariant under permutations of the configurations for the fragments, we will show that the state space of an irreducible material is a complex vector space. These states can still be compared to each other in a way that gives the state space an inner product. Such comparisons will be preserved by deterministic and reversible evolution which leads to Schroedinger equation (i.e. a unitary transformation).

@@

The overall layout of the work is a balance between the need to show that the derivation follows (which requires starting from basic principles and reaching conclusion using formal arguments) and the need to give some intuitive understanding (which often requires starting from specific cases and then generalizing).

The deductive aspect is formally organized into the following:
\begin{description}
  \item[Assumptions] these characterize the physical system we are studying and constitute the premise of our discussion. A \textbf{rationale} follows each assumption, which uses physical and sometimes philosophical arguments to motivate why (or why not) such an assumption makes sense (in a particular case).
  \item[Propositions] these characterize the mathematical properties of the physical concepts we are studying. A \textbf{justification} follows each proposition to show why such characterization follows from the physical properties of the system. While the style of a justification is formal, it is not a pure mathematical proof as it necessarily has to mix physical and mathematical arguments.

  \item[Theorems] these are pure mathematical statements that are used to simplify propositions. A mathematical \textbf{proof} follows each proposition. No mathematical breakthrough should be expected as we mostly use well known results from different fields. Theorems are grouped in the appendix.
\end{description}

As these sections are clearly demarcated, this allows us to make sure that each justification only references past propositions. It also makes it easier to read the work at the desired level of detail.

Interspersed between each step of the more formal treatment, there are numerous discussions that conceptually explain motivations and results in a more approachable way. These can use more concrete and intuitive examples without the fear of circular arguments. As both the formal and more intuitive part need to cover the same material, this will inevitably lead to some amount of repetition.

The work uses ideas and tools from a broad range of mathematical and scientific subjects: topology, measure theory, differential geometry (symplectic and Rimannian), group theory and vector spaces, statistics, information theory, Hamiltonian mechanics, Lagrangian mechanics, quantum mechanics, thermodynamics, special relativity and so on. Connecting these areas, in fact, is part of the insight given by this work. While this is a necessity, it does create a number of problems.

The first one is terminology and prerequisite knowledge. Since we touch a good number of areas, we cannot expect all readers to be equally comfortable with all the technical terms. Yet we cannot introduce all of them either. The compromise we adopted is to keep the informal discussion more accessible while keeping the formal derivation more precise. Even in the formal derivation, though, we will seldom need more than the basic definitions within a specific area and few major results.

The second problem is notation. As each field has developed its own set of conventions, the common usage is now an empty set. We ended up with a compromise that we believe strikes a good balance. Especially in the informal discussion, we crafted something that in general feels familiar to a physicist. Within formal parts, though, we may use notation and terminology more specific to the area of math being used. But then again: one would need to be familiar to that area anyway to be able to follow the details.

We have tested the prose on a number of students and colleagues to be confirm its accessibility at different levels. We hope the reader enjoys this journey and comes out with a few new insights and ideas.
